---
title: "Data Reduction"
output: html_document
date: "2025-05-31"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
MS_Table_norm = read.table("Datensätze/MS_Table.norm.csv", header=TRUE, row.names=1, sep = ",")
```

## Principal Component Analysis on Original Data

### Reduction of data-dimensionality by identifying the main axes of variability

##### Authors: Mirjam Biollaz, Jette Klempt-Gießing, Alicia Weeber

```{r}
# PCA with original values
# only Carl's code
pca = prcomp(MS_Table, center = TRUE, scale = TRUE)
summary(pca)
pca$sdev
variance = (pca$sdev)^2
prop.variance = variance / sum(variance)
names(prop.variance) = 1:length(prop.variance)
barplot(prop.variance[1:20], ylab = 'Proportion of variance') # we only plot the first 20 PCs

plot(pca$x[,1], pca$x[,2], 
     col = 'black', pch = 19,
     xlab = 'PC1', ylab = 'PC2')
```

## Separation of Normalized Data by Experimental Condition

##### Authors: Hasset Gessese

```{r}
# Number of fractions
n_fractions <- 25

# Prepare column indices
control_idx <- c()
rnase_idx <- c()

for (f in 0:(n_fractions - 1)) {
  start_col <- 6 * f
  # Control: columns 1-3 of each fraction
  control_idx <- c(control_idx, start_col + 1:3)
  # RNAse: columns 4-6 of each fraction
  rnase_idx <- c(rnase_idx, start_col + 4:6)
}

# Extract tables
MS_Control <- MS_Table_norm[, control_idx]
MS_RNase <- MS_Table_norm[, rnase_idx]
```

## Installing mentioned libraries

### The required packages for clustering analysis and visualization are acquired

##### Authors: Hasset Gessese

```{r install_all_needed, eval=FALSE}
install.packages(c("factoextra", "cluster", "pheatmap", "RColorBrewer"))
# For Elbow and Silhouette methods => library(factoextra)
# For silhouette plots => library(cluster)
# For heatmaps => library(pheatmap)
# For color palettes in plots => library(RColorBrewer)
```

## K-Means Clustering

### PCA-Based K-Means Clustering for Normalization Assessment

##### Authors: Hasset Gessese

This K-means clustering was used as a quality control check after
normalization, in order to ensure that the data structure remained
interpretable and consistent (shows seperability and stability). Because
the optimal k (k=6) from this approach did not align with our modeling
results, we did not use these clusters for biological conclusions

```{r kmeans_pca_clustering, message=FALSE, warning=FALSE}

# Perform Principal Component Analysis (PCA) on the normalized data
pca <- prcomp(MS_Table_norm, center = TRUE, scale. = TRUE)

# Calculate the proportion of variance explained by each principal component
prop.variance <- (pca$sdev)^2 / sum(pca$sdev^2)
pc1_var <- round(prop.variance[1] * 100, 1)  # Variance % explained by PC1
pc2_var <- round(prop.variance[2] * 100, 1)  # Variance % explained by PC2

# Visualize the variance proportions for the top 10 PCs with a barplot
barplot(prop.variance[1:10], main = "Variance Proportions (Top 10 PCs)", ylab = "Proportion")

# Select the first 5 principal components as features for clustering
pca_data <- pca$x[, 1:5]

# Define a custom color palette (poster colors) for cluster visualization
cluster_colors <- c(
  "#7E51A4", "#504DA4", "#485571",
  "#7396AD", "#DBCEE6", "#A9B3B5"
)

# Load required libraries for clustering visualization and analysis
library(factoextra)  # For cluster visualization and determining cluster number
library(cluster)     # For silhouette analysis

# Use the Elbow method to estimate the optimal number of clusters (k)
elbow_plot <- fviz_nbclust(pca_data, kmeans, method = "wss") +
  ggtitle("Elbow Method – Optimal Number of Clusters") +
  geom_vline(xintercept = 6, linetype = 2, color = "blue")  # Mark k=6 for reference
print(elbow_plot)

# Use the Silhouette method to validate the optimal cluster number
silhouette_plot <- fviz_nbclust(pca_data, kmeans, method = "silhouette") +
  ggtitle("Silhouette Method – Optimal Number of Clusters")
print(silhouette_plot)

# Run K-means clustering with k=6 clusters and multiple random starts for stability
set.seed(123)  # Ensure reproducibility
km <- kmeans(pca_data, centers = 6, nstart = 25)
cluster_assignment <- km$cluster  # Extract cluster membership

# Plot the first two PCs with points colored by cluster assignment
plot(
  pca$x[, 1], pca$x[, 2],
  col = cluster_colors[cluster_assignment],
  pch = 19,
  xlab = paste0("PC1 (", pc1_var, "% Variance)"),
  ylab = paste0("PC2 (", pc2_var, "% Variance)"),
  main = "K-Means Clustering on PCA (k=6)"
)

# Draw convex hulls around points of each cluster to visually group them
library(grDevices)  # For polygon drawing
for (k in 1:max(cluster_assignment)) {
  cluster_points <- pca$x[cluster_assignment == k, 1:2]
  if (nrow(cluster_points) >= 3) {  # Need at least 3 points to form a polygon
    hull_indices <- chull(cluster_points)          # Indices of convex hull points
    hull_indices <- c(hull_indices, hull_indices[1])  # Close the polygon
    polygon(
      cluster_points[hull_indices, ],
      border = cluster_colors[k],
      col = adjustcolor(cluster_colors[k], alpha.f = 0.4),  # Semi-transparent fill
      lwd = 2
    )
  }
}

# Prepare for silhouette analysis to assess cluster quality
n_obs <- nrow(pca_data)
sample_threshold <- 1000

# If dataset is large, use a random sample of 500 points for efficiency
if (n_obs > sample_threshold) {
  set.seed(42)
  sample_idx <- sample(1:n_obs, 500)
  pca_sample <- pca_data[sample_idx, ]
  cluster_sample <- cluster_assignment[sample_idx]
  dist_sample <- dist(pca_sample)  # Compute distance matrix on the sample
  sil <- silhouette(cluster_sample, dist_sample)
  message("The silhouette plot is based on a sample of 500 observations")
} else {
  # Otherwise, compute silhouette on full dataset
  dist_matrix <- dist(pca_data)
  sil <- silhouette(cluster_assignment, dist_matrix)
}

library(factoextra)  # For silhouette visualization

# Calculate average silhouette width overall and per cluster
avg_sil_width <- mean(sil[, "sil_width"])
avg_sil_per_cluster <- tapply(sil[, "sil_width"], sil[, "cluster"], mean)

# Prepare annotation text showing average silhouette widths
label_text <- paste0("Average Silhouette Width: ", round(avg_sil_width, 3), "\n",
                     paste0("Cluster ", names(avg_sil_per_cluster), ": ",
                            round(avg_sil_per_cluster, 3),
                            collapse = "\n"))

# Generate silhouette plot with cluster colors and annotated averages
sil_plot <- fviz_silhouette(sil, palette = cluster_colors) +
  ggtitle("Silhouette Plot for K-Means Clustering") +
  theme(plot.title = element_text(hjust = 0.5))

# Add annotation text to the plot
sil_plot +
  annotate("text",
           x = Inf, y = Inf,
           label = label_text,
           hjust = 1.05, vjust = 1.05,
           size = 3,
           fontface = "italic")

# Export the cluster assignments as a CSV file for downstream analysis
cluster_df <- data.frame(Protein = rownames(MS_Table_norm),
                         Cluster = cluster_assignment)

write.csv(cluster_df, "Cluster_Assignment.csv", row.names = FALSE)
```

## Visualization of a PCA defined by 5 shift categories

### PCA plot with potential biological conditions (no shift, moderate left shift, strong left shift, moderate right shift, strong right shift)

##### Authors: Alicia Weeber

```{r}
# Your PCA object
pca = prcomp(MS_Table_norm, center = TRUE, scale. = TRUE)

# PCA scores (coordinates of the proteins on the PCs)
scores = pca$x

# Get the shift categories from the results table
shift_kategorie = df_ergebnisse[,14]  # Column 14

# Check that the order of categories matches the PCA scores
# If MS_Table_norm and df_ergebnisse have the same row order, it's fine.
# If not, you need to make sure they are matched properly!

# Convert categories to a factor (optionally with defined order)
shift_kategorie = factor(shift_kategorie, levels = c("kein_shift",
                                                     "moderater_left_shift",
                                                     "starker_left_shift",
                                                     "moderater_right_shift",
                                                     "starker_right_shift"))

# Define colors for the categories
farben = c("kein_shift"="grey50",
           "moderater_left_shift"="orange",
           "starker_left_shift"="red",
           "moderater_right_shift"="skyblue",
           "starker_right_shift"="blue")

# Layout: 1 row, 2 columns (increase space for legend)
layout(matrix(c(1,2), 1, 2), widths=c(3.5,1.5))  # Adjusted layout to avoid overlap

# PCA plot
par(mar=c(5,4,4,2))  # Slightly larger right margin for spacing
plot(scores[,1], scores[,2],
     col=farben[shift_kategorie],
     pch=19,
     xlab="PC1", ylab="PC2",
     main="PCA: Coloring by Shift Category")

# Legend in separate panel
par(mar=c(2,0,2,0))  # Minimal vertical padding
plot.new()           # empty plot window
legend("center", legend=levels(shift_kategorie),
       col=farben, pch=19, title="Shift Category", cex=0.9)
```

```{r}
df_ergebnisse
```
