---
title: "Untitled"
output: html_document
date: "2025-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
MS_Table_norm = read.table("Datensätze/MS_Table.norm.csv", header=TRUE, row.names=1, sep = ",")
```

## General goal of data modelling:
The general goal of the data modelling is on the one hand to evaluate how well the classification of RBPs oder NO_RBPs based on the shift score is. On the other hand the main goal is tto further analyze and characterize the identified RNA-dependent proteins by their biophysical characteristics which is done by performing linear and logistic regression analyses.

# Loading the data from Rdeep and creating a table of RBP listed proteins
```{r}
raw <- readLines("Datensätze/Table_HS_RBP.txt")
raw_data <- raw[7:length(raw)]
clean_data <- gsub('\\"', '', raw_data)
writeLines(clean_data, "cleaned_table.txt")
Table_HS_RBP = read.delim("cleaned_table.txt",
                    sep = "\t",
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    check.names = FALSE)
```

# Loading the data from Rdeep and creating a table of proteins listed as NO_RBP
```{r}
raw <- readLines("Datensätze/Table_HS_Non_RBP.txt")
raw_data <- raw[7:length(raw)]
clean_data <- gsub('\\"', '', raw_data)
writeLines(clean_data, "cleaned_table.txt")
Table_HS_NO_RBP = read.delim("cleaned_table.txt",
                    sep = "\t",
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    check.names = FALSE)
```

```{r}
# Combining the tables "RBP_Table" and "No_RBP_Table" 
rbp_cols <- colnames(Table_HS_RBP)

#selecting columns excisting in both tables
existing_no_rbp_cols <- intersect(rbp_cols, colnames(Table_HS_NO_RBP))

# NO_RBP_Table with the same columns
No_RBP_Table_aligned <- Table_HS_NO_RBP[, existing_no_rbp_cols]

# Adding missing columns with NA
missing_cols <- setdiff(rbp_cols, existing_no_rbp_cols)
for (col in missing_cols) {
  No_RBP_Table_aligned[[col]] <- NA
}
# Same order of the columns in both tables
No_RBP_Table_aligned <- No_RBP_Table_aligned[, rbp_cols]

# Categorizing the proteins based on their original table
Table_HS_RBP$Kategorie <- "RBP"
No_RBP_Table_aligned$Kategorie <- "NO_RBP"

# Keeping the same entries a sin Table_MS_norm
RBP_Table <- Table_HS_RBP[Table_HS_RBP$Entry_Name %in% rownames(MS_Table_norm), ]
No_RBP_Table <- No_RBP_Table_aligned[No_RBP_Table_aligned$Entry_Name %in% rownames(MS_Table_norm), ]

# Combinig both tables for further analysis
combined_table <- rbind(RBP_Table, No_RBP_Table)

```

```{r}
# Loading data from Uniprot for more characteristics of the proteins
library(readr)
library(dplyr)

df_uniprot <- read_tsv("Datensätze/Uniprot_infos2.tsv", comment = "#", quote = "")

df_uniprot_counts <- df_uniprot %>%
  mutate(
    Zinc_Count      = ifelse(is.na(`Zinc finger`)      | `Zinc finger` == "",      NA, lengths(strsplit(`Zinc finger`, ";\\s*"))),
    Coil_Count      = ifelse(is.na(`Coiled coil`)      | `Coiled coil` == "",      NA, lengths(strsplit(`Coiled coil`, ";\\s*"))),
    Signal_Count    = ifelse(is.na(`Signal peptide`)   | `Signal peptide` == "",   NA, lengths(strsplit(`Signal peptide`, ";\\s*"))),
    Glyco_Count     = ifelse(is.na(`Glycosylation`)    | `Glycosylation` == "",    NA, lengths(strsplit(`Glycosylation`, ";\\s*"))),
    Disulfide_Count = ifelse(is.na(`Disulfide bond`)   | `Disulfide bond` == "",   NA, lengths(strsplit(`Disulfide bond`, ";\\s*"))),
    TM_Count        = ifelse(is.na(`Transmembrane`)    | `Transmembrane` == "",    NA, lengths(strsplit(`Transmembrane`, ";\\s*")))
  ) %>%
  select(Entry, `Entry Name`, Zinc_Count, Coil_Count, TM_Count)

# Comining the new characterisitics
combined_table <- combined_table %>%
  left_join(df_uniprot_counts, by = c("Entry_Name" = "Entry Name"))
```

```{r}
# adding the biophysical traits of the proteins in df_ergebnisse for the regression analysis
phys_traits = combined_table[, c("Entry_Name", "Mass_kDa", "Length_AA", "pI", "Listing_Count","Zinc_Count", "Coil_Count", "RBP2GO_Score", "TM_Count")]


df_ergebnisse_phys = merge(df_ergebnisse, phys_traits,
                   by.x = "Protein", 
                   by.y = "Entry_Name", all = FALSE)

df_ergebnisse_phys$RBP_Status = ifelse(df_ergebnisse_phys$Listing_Count >= 5, "RBP", "NO_RBP")

#Einteilung der Proteine ein RBP/ kein RBP auf Basis des SHift_Scores
df_ergebnisse_phys$RBP_nach_Score <- ifelse(df_ergebnisse_phys$Shift_Kategorie == "no_shift", "NO_RBP", "RBP")
df_ergebnisse_phys$RBP_nach_Score <- factor(df_ergebnisse_phys$RBP_nach_Score, levels = c("NO_RBP", "RBP"))
```


# Evaluating the RBP classification based on the shift category and shift score by comparing it to the data from Rdeep
```{r}
# Vergleich Kategorisierung nach Shift_Score vs. nach Datenbank
conf_matrix = table(df_ergebnisse_phys$RBP_Status, df_ergebnisse_phys$RBP_nach_Score)
colnames(conf_matrix) <- c("Vorhergesagt_NO_RBP", "Vorhergesagt_RBP")
rownames(conf_matrix) <- c("Tatsächlich_NO_RBP", "Tatsächlich_RBP")
conf_matrix

TN <- 4578
FP <- 800
FN <- 985
TP <- 697

# Funktion zur Berechnung der Klassifikationsmetriken
calc_metrics <- function(TP, TN, FP, FN) {
  Gesamttrefferquote <- (TP + TN) / (TP + TN + FP + FN)
  PPW <- TP / (TP + FP)
  Sensitivität <- TP / (TP + FN)  # Sensitivität
  Spezifität <- TN / (TN + FP)
  f1_score <- 2 * (PPW * Sensitivität) / (PPW + Sensitivität)
  
  # Ergebnisse als Dataframe
  metrics_df <- data.frame(
    Metric = c("Gesamttrefferquote", "PPW", "Sensitivität", "Spezifität", "F1-Score"),
    Value = c(Gesamttrefferquote, PPW, Sensitivität, Spezifität, f1_score)
  )
  return(metrics_df)
}

Testwerte <- calc_metrics(TP, TN, FP, FN)
print(Testwerte)
```

# Linear regression analysis: Isoelectric point and mass as predictors for the shift_score
Several linear regression analyses were performed to find correlations between different biophysical characteristics of the proteins and the Shift score. Many of those analyses are not shown anymore because they all did not identify significant predictors for the shift score. 
One linear regression analysis is performed with the isoelectric point and mass of the proteins as predictors for their Shift_Score. The goal of this regression analysis is to find out whether those two biophysical traits have a significant impact on the Shift_Score and therefore on the categorization of the proteins as RBP or NO_RBP.
```{r}
model_linear = lm(Shift_Score ~ pI + Mass_kDa, data = df_ergebnisse_phys)
summary(model_linear)
```

Both, the isoelectric point as well as the mass have a significant correlation with the Shift_Score because the p-value for both coefficients is significant. The correlation between those two traits and the Shift_Score is visualized below.

```{r}
df_plot <- df_ergebnisse_phys[complete.cases(df_ergebnisse_phys[, c("RBP_nach_Score", "Mass_kDa", "pI")]), ]
df_plot <- df_plot[df_plot$Mass_kDa < 2000, ]
df_plot$predicted_score <- predict(model_linear, newdata = df_plot, type = "response")

ggplot(df_plot, aes(x = Mass_kDa, y = pI, color = predicted_score)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "Shift_Score based on mass and pI",
       x = "Masse", y = "Isoelektrischer Punkt (pI)",
       color = "Score") +
  theme_minimal()
```
```{r}
# Dichteverteilungen
ggplot(df_ergebnisse_phys, aes(x = Mass_kDa, y = pI)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  facet_wrap(~RBP_nach_Score) +
  labs(x = "Masse", y = "pI", title = "2D-Dichte von Masse & pI nach Shift Kategorie") +
  theme_minimal()
```


```{r}
# pI und Masse isoliert betrachten als Prädiktor für den Shift Score
# Zusammenhang zwischen Veränderung von pI bzw. Masse und dem shift score
# Fläche ist 95% Konfidenzintervall

install.packages("effects")
library(effects)

plot(allEffects(model_linear), main = "Teil-Effekte der Prädiktoren auf Shift Score")
```

The R-squared value of the regression analysis is 0.03363 indicating that the isoelectric point and mass alone are not sufficient enough to predict the Shift_Score of the proteins. This result is visualized below. Both traits have a linear correlation with the shift score but there is a big deviation noticeable showing that these traits alone do not eexplain the Shift_Score. 

```{r}
ggplot(df_plot, aes(x = pI, y = Shift_Score, color = Shift_Kategorie)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Influence of pI on shift score",
       x = "Isoelekctric point (pI)",
       y = "Shift score") +
  theme_minimal()
ggplot(df_plot, aes(x = Mass_kDa, y = Shift_Score, color = Shift_Kategorie)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Influence of mass on shift score",
       x = "Mass (kDa)",
       y = "Shift score") +
  theme_minimal()
```

```{r}
df_ergebnisse_phys$fitted_shift = predict(model_linear, newdata = df_ergebnisse_phys)
```

```{r}
summary(df_ergebnisse_phys$fitted_shift)
range(df_ergebnisse_phys$fitted_shift, na.rm = TRUE)
```


```{r}
# Vergleich vorrherrgesagter Score durch das Modell und dem bestimmten Shift_Score
# gibt linearen Zusammenhang, aber viel Streuung
# Prädiktoren aus dem Modell allein erklären den Shift nicht

plot(df_ergebnisse_phys$Shift_Score, df_ergebnisse_phys$fittet_shift,
     xlab = "Gemessener Shift Score",
     ylab = "Vorhergesagter (fitted) Shift Score",
     main = "Vergleich gemessener vs. vorhergesagter Shift Scores")
abline(a=0, b=1, col="red")  # ideale Gerade y=x
# gemessener und vorausgesagter shift stimmen nicht wirklch überein - Info über die eigenschaften reicht eigentlich nicht aus, um shift_score zu bestimmen

plot(df_ergebnisse_phys$Shift_Score, df_ergebnisse_phys$fitted_shift,
     xlab = "Gemessener Shift Score", ylab = "Vorhergesagter Shift Score",
     main = "Vergleich gemessener vs. vorhergesagter Shift Scores")
abline(a=0, b=1, col="red")

```

# Data reduction and cluster analysis

Performing a data reduction and cluster analysis based on the coefficients of the linear regression analysis, the goal is to find out the number of clusters and comparing those with the assignment of the proteins in different shift categories.

```{r}
# Choosing the variables for the clusteriing
vars = c("Mass_kDa", "pI", "Shift_Score")
df_clust = df_ergebnisse_phys[complete.cases(df_ergebnisse_phys[, vars]), ]

# Scaling to prepare for clustering
scaled_data = scale(df_clust[, vars])

# k-means clustering 
set.seed(42)  # für Reproduzierbarkeit
kmeans_result = kmeans(scaled_data, centers = 4)

# Saving cluster assignment
df_clust$regression_cluster = factor(kmeans_result$cluster)

```

```{r}
# PCA to perform a data reduction based on the proteins with selected variables for the clustering 
pca = prcomp(scaled_data, scale. = TRUE)  

# Data for plotting
plot_df = data.frame(PC1 = pca$x[,1],
                      PC2 = pca$x[,2],
                      Cluster = df_clust$regression_cluster)


# Plot
library(ggplot2)

ggplot(plot_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Clusteriing based on regression analysis",
       x = paste0("PC1 (", round(summary(pca)$importance[2,1]*100,1), "%)"),
       y = paste0("PC2 (", round(summary(pca)$importance[2,2]*100,1), "%)")) +
  scale_color_brewer(palette = "Set2")


```
This clustering contains an outlier that skews the clustering plot so it has to be removed. The proteins are clustere again and the silhouette plot shows which the best number of clusters in this analysis is.
```{r}
# Detecting the outlier and removing it
outlier_index <- which.max(abs(plot_df$PC2))  
df_clust_no_outlier <- df_clust[-outlier_index, ]
scaled_data_no_outlier <- scale(df_clust_no_outlier[, vars])
set.seed(42)
kmeans_result <- kmeans(scaled_data_no_outlier, centers = 4)

# New cluster assignment
df_clust_no_outlier$regression_cluster <- factor(kmeans_result$cluster)
pca_no_outlier <- prcomp(scaled_data_no_outlier, scale. = FALSE)

plot_df_clean <- data.frame(
  PC1 = pca_no_outlier$x[,1],
  PC2 = pca_no_outlier$x[,2],
  Cluster = df_clust_no_outlier$regression_cluster,
  Protein = df_clust_no_outlier$Protein  
)

# Merging of shift category by protein name in the dataframe for plotting
plot_df_clean <- merge(plot_df_clean,
                       df_ergebnisse_phys[, c("Protein", "Shift_Kategorie")],
                       by = "Protein", all.x = TRUE)

library(ggplot2)

ggplot(plot_df_clean, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "Clustering by pI, mass and shift score ")

# Silhouette plot for the clustering
library(cluster)
# distance matrix
dists <- dist(scaled_data_no_outlier)

# Silhouette with k-means assignments
sil <- silhouette(as.numeric(df_clust_no_outlier$regression_cluster), dists)
plot(sil, main = "Silhouette-Plot for K-Means-Clustering (k = 4)", col = 2:7, border = NA)
```
Combining the clustering information with the assignment of each protein to a shift category, the goal is to find out if there is a visible structure in the clustering based on the information of the shiift category. The result shows that clustering the proteins by pI, mass and Shift Score does not separate the proteins based on their shift category meaning that it was succesful to visualize that pI and mass explain some of variance of the shift score. Therefore the result is that especially left shift proteins have a different isoelectric point than proteins with no shift.
```{r}
# Most abundant shift categroy in each cluster
library(dplyr)
# absolute values
summary_per_cluster <- df_clust_no_outlier %>%
  group_by(regression_cluster, Shift_Kategorie) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))

#relative values
summary_per_cluster_rel <- summary_per_cluster %>%
  group_by(regression_cluster) %>%
  mutate(
    Anteil = n / sum(n)
  ) %>%
  ungroup()

# Proportion of RBP and NO_RBP in each cluster  
summary_per_cluster_RBP <- df_clust_no_outlier %>%
  group_by(regression_cluster, RBP_nach_Score) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(n))
summary_per_cluster_rel_RBP <- summary_per_cluster_RBP %>%
  group_by(regression_cluster, RBP_nach_Score) %>%
  mutate(
    Anteil = n / sum(n)
  ) %>%
  ungroup()
```

```{r}
# Statistical test to see if the clusters distinguish in shift category
# Contingency table
table_shift_cluster <- table(df_clust_no_outlier$regression_cluster, df_clust_no_outlier$Shift_Kategorie)

# Chi²-test for shift categories
chisq.test(table_shift_cluster)
# Verteilung der shift Kategorien unterscheidet sich signifikant zwischen den Clustern 

# Chi²-test for RBP classification in clusters
# ebenfalls signifikant
table_shift_cluster_RBP <- table(df_clust_no_outlier$regression_cluster, df_clust_no_outlier$RBP_nach_Score)
chisq.test(table_shift_cluster_RBP)
```
Both, the shift categories and RBP classification distinguish in between the clusters. These results are visualized below by coloring the proteins by shift category.
```{r}
ggplot(plot_df_clean, aes(x = PC1, y = PC2, color = Shift_Kategorie)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: Clustering by pI, mass and shift score") +
  theme_minimal()

```
The distribution of each shift category in each cluster is visualzed below.
```{r}
ggplot(df_clust_no_outlier, aes(x = Shift_Kategorie, y = pI, fill = Shift_Kategorie)) +
  geom_boxplot() +
  facet_wrap(~ regression_cluster) +
  theme_minimal() +
  labs(title = "pI-values pro shift-category in each cluster")
```
To show the difference in the isoelectric point in each shift category, a violin plot is rendered.
```{r}
library(ggplot2)

ggplot(df_clust, aes(x = Shift_Kategorie, y = pI)) +
  geom_violin(fill = "skyblue", alpha = 0.4) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  theme_minimal() +
  labs(title = "Isoelectric point in each shift category", x = "Shift category", y = "pI")

```
Below there is a density distribution of the isloelectric point for the two ctegories "RBP" and "NO_RBP". It is another visualization to show the difference in the isoelectric point for the different categories.
```{r}
ggplot(df_ergebnisse_phys, aes(x = pI, fill = RBP_nach_Score)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "pI-Verteilung bei RBP und Nicht-RBP", x = "pI", y = "Dichte")
```
The violin plot for the log10(mass) in each shift category shows that there is not a significant difference visible. Without logarithmizing the mass the violin plots were not visible.
```{r}
library(ggplot2)
ggplot(df_clust, aes(x = Shift_Kategorie, y = log10(Mass_kDa))) +
  geom_violin(fill = "skyblue", alpha = 0.4) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  theme_minimal() +
  labs(
    title = "Log10(Mass) in each shift category",
    x = "Shift-category",
    y = expression(log[10]*"(Mass in kDa)")
  )

```


# Logistic regression analysis: Correlation between isoelectric point and Zinc motive with RBP classification
Finding 570 proteins that have an information about a zinc finger motive and are also in the dataset that this analysis is based on led to the goal of searching for a correlation between a zinc finger motive and the RBP classification. A linear regression analysis with the zinc finger motive was not successful so that a logistic regression is performed with the target variable of RBP classification in RBP oder NO_RBP instead of the shift score in the linear regression analysis. Because the isoelectric point was already identified as influencial on both the shift score and the RBP classification, the pI is also included in the logistic regression analysis.
```{r}
df_ergebnisse_phys$RBP_nach_Score = factor(df_ergebnisse_phys$RBP_nach_Score, levels = c("NO_RBP", "RBP"))
model_logit = glm(RBP_nach_Score ~ Zinc_Count + pI,
                   data = df_ergebnisse_phys,
                   family = "binomial")
summary(model_logit)
df_pred <- df_ergebnisse_phys[complete.cases(df_ergebnisse_phys[, c("Zinc_Count", "pI")]), ]
df_pred$fitted_Status <- predict(model_logit, newdata = df_pred, type = "response")
```

```{r}
# Wie gut stimmt das Modell mit der Klassifikation als RBP/ NO_RBP überein -> statistisch überprüfen
df_pred$fitted_Status <- ifelse(df_pred$fitted_Status > 0.5, "RBP", "NO_RBP")
df_pred$fitted_Status <- factor(df_pred$fitted_Status, levels = c("NO_RBP", "RBP"))
complete_idx <- complete.cases(df_ergebnisse_phys[, c("Zinc_Count", "pI")]) # Nur die Reihen ohne NA Werte 
df_complete <- df_ergebnisse_phys[complete_idx, ] # als Datenframe

# Vorhersagen durch das logistische Modell
df_complete$fitted_Status <- predict(model_logit, newdata = df_complete, type = "response")

# bei welcher Wahrscheinlichkeit wird als RBP klassifiziert (benötigt für den Vergleich mit RBP_nach_Score bzw mit RBP_Status)
df_complete$fitted_Status <- ifelse(df_complete$fitted_Status > 0.5, "RBP", "NO_RBP")
df_complete$fitted_Status <- factor(df_complete$fitted_Status, levels = c("NO_RBP", "RBP"))

conf_matrix <- table(
  True = df_complete$RBP_nach_Score,
  Predicted = df_complete$fitted_Status
)

colnames(conf_matrix) <- c("Vom Modell vorhergesagt: NO_RBP", "Vom Modell vorhergesagt: RBP")
rownames(conf_matrix) <- c("Klassifikation: NO_RBP", "Klassifikation: RBP")

print(conf_matrix)

TN <- 377 
FP <- 17
FN <- 143
TP <- 32

# Funktion zur Berechnung der Klassifikationsmetriken
calc_metrics <- function(TP, TN, FP, FN) {
  Gesamttrefferquote <- (TP + TN) / (TP + TN + FP + FN)
  PPW <- TP / (TP + FP)
  Sensitivität <- TP / (TP + FN)  # Sensitivität
  Spezifität <- TN / (TN + FP)
  f1_score <- 2 * (PPW * Sensitivität) / (PPW + Sensitivität)
  
  # Ergebnisse als Dataframe
  metrics_df <- data.frame(
    Metric = c("Gesamttrefferquote", "PPW", "Sensitivität", "Spezifität", "F1-Score"),
    Value = c(Gesamttrefferquote, PPW, Sensitivität, Spezifität, f1_score)
  )
  return(metrics_df)
}

Testwerte <- calc_metrics(TP, TN, FP, FN)
print(Testwerte)

```

```{r}
library(dplyr)

df_ergebnisse_phys %>%
  group_by(RBP_Status) %>%
  summarise(
    mittlerer_pI = mean(pI, na.rm = TRUE),
    median_pI = median(pI, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(desc(mittlerer_pI))

```

```{r}
# Extracting the amount of RBPs with zinc motive
amount_RBP <- sum(df_ergebnisse_phys$RBP_nach_Score == "RBP")
amount_RBP_zinc <- sum(df_ergebnisse_phys$Zinc_Count > 0 & df_ergebnisse_phys$RBP_nach_Score == "RBP", na.rm = TRUE)

# Extracting the amount of NO_RBPs with zinc motive
amount_NO_RBP <- sum(df_ergebnisse_phys$RBP_nach_Score == "NO_RBP")
amount_NO_RBP_zinc <- sum(df_ergebnisse_phys$Zinc_Count > 0 & df_ergebnisse_phys$RBP_nach_Score == "NO_RBP", na.rm = TRUE)

# Percentage of both
prop_RBP_zinc <- amount_RBP_zinc / amount_RBP * 100
prop_NO_RBP_zinc <- amount_NO_RBP_zinc / amount_NO_RBP * 100
table(prop_RBP_zinc,prop_NO_RBP_zinc)
```


```{r}
t.test(Zinc_Count ~ RBP_nach_Score, data = df_ergebnisse_phys)
library(ggplot2)

ggplot(df_ergebnisse_phys, aes(x = RBP_Status, y = Zinc_Count, fill = RBP_Status)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  labs(title = "Vergleich des isoelektrischen Punkts (pI) zwischen RBP und NO_RBP",
       x = "Protein-Typ", y = "pI") +
  theme_minimal()

```

# Visualisierung der logistischen Regression
```{r}
# Dichte von pI und Zinc_Count nach RBP Score
library(ggplot2)

# Zinc_Count
ggplot(df_ergebnisse_phys, aes(x = Zinc_Count, fill = RBP_nach_Score)) +
  geom_density(alpha = 0.5) +
  labs(title = "Dichteverteilung von Zinc_Count nach RBP Score",
       x = "Zinc_Count", fill = "RBP_nach_Score") +
  theme_minimal()

# pI
ggplot(df_ergebnisse_phys, aes(x = pI, fill = RBP_nach_Score)) +
  geom_density(alpha = 0.5) +
  labs(title = "Dichteverteilung von pI nach RBP Score",
       x = "Isoelektrischer Punkt (pI)", fill = "RBP_nach_Score") +
  theme_minimal()
```

```{r}
# 2D Dichteverteilung -> eventuell noch besser bei verbesserter Funktion??
df_plot <- df_ergebnisse_phys[complete.cases(df_ergebnisse_phys[, c("RBP_nach_Score", "Zinc_Count", "pI")]), ]

ggplot(df_plot, aes(x = Zinc_Count, y = pI)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  facet_wrap(~RBP_nach_Score) +
  scale_fill_viridis_c() +
  labs(title = "2D-Dichteverteilung von Zinc_Count und pI nach RBP_Score",
       x = "Zinc_Count", y = "pI") +
  theme_minimal()
```

```{r}
# vorhergesagte Wahrscheinlichkeit (auf Basis der Regreession) für RBP anhand von pI und Zink
df_plot <- df_ergebnisse_phys[complete.cases(df_ergebnisse_phys[, c("RBP_nach_Score", "Zinc_Count", "pI")]), ]
df_plot$predicted_prob <- predict(model_logit, type = "response")

ggplot(df_plot, aes(x = Zinc_Count, y = pI, color = predicted_prob)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_gradient(low = "#EDE6F2", high = "#394041") +
  labs(title = "Vorhergesagte Wahrscheinlichkeit für RBP",
       x = "Zink-Finger Anzahl", y = "Isoelektrischer Punkt (pI)",
       color = "P(RBP)") +
  theme_minimal()
```

```{r}
# Zusammenhang pI und Zink für RBP nach Score
ggplot(df_ergebnisse_phys, aes(x = Zinc_Count, y = pI, color = RBP_nach_Score)) +
  geom_point(alpha = 0.6) +
  labs(x = "Zink-Finger-Anzahl", y = "Isoelektrischer Punkt (pI)", color = "RBP nach Score") +
  theme_minimal()
```

```{r}
# Zusammenhang pI und Zink für RBP Status (Datenbank) 
# Regressionsmodell stimmt noch nicht sehr hiermit überein -> eventuell nach verbesserter Funktion?
ggplot(df_ergebnisse_phys, aes(x = Zinc_Count, y = pI, color = RBP_Status)) +
  geom_point(alpha = 0.6) +
  labs(x = "Zink-Finger-Anzahl", y = "Isoelektrischer Punkt (pI)", color = "RBP Status") +
  theme_minimal()
```

# Visualisieruung der Ergebnisse der Shift und RBP Klassifizierung

```{r}
library(dplyr)
library(ggplot2)

# RBP-Status zuordnen
df_ergebnisse_phys <- df_ergebnisse_phys %>%
  mutate(RBP_Gruppe = ifelse(Shift_Kategorie == "kein_shift", "non-RBP", "RBP"))

# Summen berechnen pro RBP_Gruppe + Shift_Kategorie
df_summary <- df_ergebnisse_phys %>%
  group_by(RBP_Gruppe, Shift_Kategorie) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    # Kategorie-Label mit Anzahl für die Legende
    Kategorie_mit_n = paste0(Shift_Kategorie, " (", n, ")")
  )

ggplot(df_summary, aes(x = RBP_Gruppe, y = n, fill = Kategorie_mit_n)) +
  geom_col(position = "stack") +
  # Kein geom_text() → keine Zahlen in den Balken
  scale_fill_manual(
    values = setNames(
      c("#A786C2", "#8684C2", "#7396AD", "#728083","#DBCEE6"), 
      df_summary$Kategorie_mit_n
    )
  ) +
  labs(title = "Anzahl an Proteinen pro Shift-Kategorie",
       x = "Protein-Typ", y = "Anzahl Proteine", fill = "Shift-Kategorie (n)") +
  theme_minimal(base_size = 13)

```



