---
title: "Datenset"
output: html_document
date: "2025-04-30"
---

```{r setup, include=FALSE}
# Setup chunk: set default chunk options
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load data table
MS_Table = read.table("Datensätze/RDeeP_HeLa_Interphase.csv", header=TRUE, row.names=1, sep = ";")
```

```{r}
Fraktionsnamen = c('Fraction1_Ctrl', 'Fraction1_RNAse', 'Fraction2_Ctrl', 'Fraction2_RNAse','Fraction3_Ctrl', 'Fraction3_RNAse','Fraction4_Ctrl', 'Fraction4_RNAse','Fraction5_Ctrl', 'Fraction5_RNAse','Fraction6_Ctrl', 'Fraction6_RNAse','Fraction7_Ctrl', 'Fraction7_RNAse','Fraction8_Ctrl', 'Fraction8_RNAse','Fraction9_Ctrl', 'Fraction9_RNAse','Fraction10_Ctrl', 'Fraction10_RNAse','Fraction11_Ctrl', 'Fraction11_RNAse','Fraction12_Ctrl', 'Fraction12_RNAse','Fraction13_Ctrl', 'Fraction13_RNAse','Fraction14_Ctrl', 'Fraction14_RNAse','Fraction15_Ctrl', 'Fraction15_RNAse','Fraction16_Ctrl', 'Fraction16_RNAse','Fraction17_Ctrl', 'Fraction17_RNAse','Fraction18_Ctrl', 'Fraction18_RNAse','Fraction19_Ctrl', 'Fraction19_RNAse','Fraction20_Ctrl', 'Fraction20_RNAse','Fraction21_Ctrl', 'Fraction21_RNAse','Fraction22_Ctrl', 'Fraction22_RNAse','Fraction23_Ctrl', 'Fraction23_RNAse','Fraction24_Ctrl', 'Fraction24_RNAse','Fraction25_Ctrl', 'Fraction25_RNAse')
```

```{r}
meantabellen = c('mean_Fraction1_Ctrl', 'mean_Fraction1_RNAse', 'mean_Fraction2_Ctrl', 'mean_Fraction2_RNAse','mean_Fraction3_Ctrl', 'mean_Fraction3_RNAse','mean_Fraction4_Ctrl', 'mean_Fraction4_RNAse','mean_Fraction5_Ctrl', 'mean_Fraction5_RNAse','mean_Fraction6_Ctrl', 'mean_Fraction6_RNAse','mean_Fraction7_Ctrl', 'mean_Fraction7_RNAse','mean_Fraction8_Ctrl', 'mean_Fraction8_RNAse','mean_Fraction9_Ctrl', 'mean_Fraction9_RNAse','mean_Fraction10_Ctrl', 'mean_Fraction10_RNAse','mean_Fraction11_Ctrl', 'mean_Fraction11_RNAse','mean_Fraction12_Ctrl', 'mean_Fraction12_RNAse','mean_Fraction13_Ctrl', 'mean_Fraction13_RNAse','mean_Fraction14_Ctrl', 'mean_Fraction14_RNAse','mean_Fraction15_Ctrl', 'mean_Fraction15_RNAse','mean_Fraction16_Ctrl', 'mean_Fraction16_RNAse','mean_Fraction17_Ctrl', 'mean_Fraction17_RNAse','mean_Fraction18_Ctrl', 'mean_Fraction18_RNAse','mean_Fraction19_Ctrl', 'mean_Fraction19_RNAse','mean_Fraction20_Ctrl', 'mean_Fraction20_RNAse','mean_Fraction21_Ctrl', 'mean_Fraction21_RNAse','mean_Fraction22_Ctrl', 'mean_Fraction22_RNAse','mean_Fraction23_Ctrl', 'mean_Fraction23_RNAse','mean_Fraction24_Ctrl', 'mean_Fraction24_RNAse','mean_Fraction25_Ctrl', 'mean_Fraction25_RNAse')
```

```{r}
Replikatsnamen = c('Ctrl1_norm_mittel','Ctrl2_norm_mittel','Ctrl3_norm_mittel','RNAse1_norm_mittel','RNAse2_norm_mittel','RNAse3_norm_mittel')
```

```{r}
SW_Replikatsnamen = c('SW_Ctrl1','SW_Ctrl2','SW_Ctrl3','SW_RNAse1','SW_RNAse2','SW_RNAse3')
```

```{r}
SW_Replikatsnamen_norm = c('SW_Ctrl1_norm','SW_Ctrl2_norm','SW_Ctrl3_norm','SW_RNAse1_norm','SW_RNAse2_norm','SW_RNAse3_norm')
```

```{r}
Treatmentnamen = c('Ctrl1_norm','Ctrl2_norm','Ctrl3_norm','RNAse1_norm','RNAse2_norm','RNAse3_norm')
```

# Description of the Data

### Dimensions of the table:

```{r}
dim(MS_Table)
```

### colnames(MS_Table)

```{r}
colnames(MS_Table)
```

### rownames(MS_Table)

```{r}
rownames(MS_Table)
```

### Number of missing values

```{r}
sum(is.na(MS_Table))  # Total number of NAs

sum(rowSums(MS_Table) == 0)  # Number of rows with only zeros
which(rowSums(MS_Table) == 0)  # Index of those rows → must be removed

```

### Check if all values are numeric

```{r}
sum(sapply(MS_Table, is.numeric))
# This returns 150 → means all columns are numeric
```

### Remove FHOD3_HUMAN

```{r}
MS_Table = MS_Table[setdiff(rownames(MS_Table), 'FHOD3_HUMAN'),]
```

# Normalization

```{r}
# Create subtables
for (i in 1:50) {
  spalten_index = ((i-1) * 3 + 1):(i * 3)  # Each loop processes 3 columns together
  teilmatrix = MS_Table[,spalten_index]
  
  assign(Fraktionsnamen[i], teilmatrix)  # Automatically assigns the correct name
}
```

```{r}
# Column means for each submatrix as vector
for (i in 1:50) {
  spalten_index = ((i-1) * 3 + 1):(i * 3)  # Each loop processes 3 columns together
  teilmatrix = MS_Table[,spalten_index]
  
  Mittelwerte = colMeans(teilmatrix)
  
  assign(paste0('mean_', Fraktionsnamen[i]), Mittelwerte)  # Automatically assigns correct name
}
```

```{r}
# Normalization factor per fraction
normf.fraktion = c()
for (i in 1:50) {
    
  # Calculate pairwise differences of the individual means
  abstand_12 = abs(get(meantabellen[i])[1] - get(meantabellen[i])[2])
  abstand_13 = abs(get(meantabellen[i])[1] - get(meantabellen[i])[3])
  abstand_23 = abs(get(meantabellen[i])[2] - get(meantabellen[i])[3])
  
  # Determine the smallest difference and compute the mean of the closest pair
  if (abstand_12 <= abstand_13 && abstand_12 <= abstand_23) {
    normf.fraktion[i] = mean(c(get(meantabellen[i])[1], get(meantabellen[i])[2]))
  } else if (abstand_13 <= abstand_12 && abstand_13 <= abstand_23) {
    normf.fraktion[i] = mean(c(get(meantabellen[i])[1], get(meantabellen[i])[3]))
  } else {
    normf.fraktion[i] = mean(c(get(meantabellen[i])[2], get(meantabellen[i])[3]))
  }
}
normf.fraktion
```

```{r}
# Normalization factor for each replicate
# Divide the normalization factor per fraction by each replicate’s mean
normf.rep = c()
for (i in 1:50) {
  for (j in 1:3) {
    normf.rep = c(normf.rep, normf.fraktion[i] / get(meantabellen[i])[j])
  }
}
normf.rep
```

```{r}
# Normalize the data to the mean
# Apply replicate-specific normalization factors to each corresponding replicate
MS_Table_norm_mittel = as.data.frame(matrix(NA, nrow = nrow(MS_Table), ncol = 150))
for (i in 1:150) {
  MS_Table_norm_mittel[,i] = normf.rep[i] * MS_Table[,i]
}
colnames(MS_Table_norm_mittel) = colnames(MS_Table)
rownames(MS_Table_norm_mittel) = rownames(MS_Table)

MS_Table_norm_mittel
```

```{r}
# Normalize the data to 100 using a sliding window
# Create subtables for individual replicates

for (i in 1:6) {
  spalten_index = seq(from = i, to = 150, by = 6)
  teilmatrix = MS_Table_norm_mittel[, spalten_index]
  
  assign(Replikatsnamen[i], teilmatrix)
}
```

```{r}
# Apply sliding window
for (i in 1:6) {
  
  teilmatrix = data.frame(
    get(Replikatsnamen[i])[1], 
    (get(Replikatsnamen[i])[1:23] + get(Replikatsnamen[i])[2:24] + get(Replikatsnamen[i])[3:25]) / 3, 
    get(Replikatsnamen[i])[25]
  )
  
  assign(SW_Replikatsnamen[i], teilmatrix)  # Automatically assigns the correct name
}
```


```{r}
# Normalize sliding window output to 100%
for (i in 1:6) {
  teilmatrix = (get(SW_Replikatsnamen[i])) * 100 / rowSums(get(SW_Replikatsnamen[i]))
  teilmatrix[is.na(teilmatrix)] = 0  # Handles cases where a fraction's total is 0 → prevents division
  
  assign(SW_Replikatsnamen_norm[i], teilmatrix)  # Automatically assigns the correct name
}
```

```{r}
# Merge into one final table

matrices <- list(SW_Ctrl1_norm, SW_Ctrl2_norm, SW_Ctrl3_norm, SW_RNAse1_norm, SW_RNAse2_norm, SW_RNAse3_norm)

MS_Table_norm = c()

for (i in 1:25) {
  for (j in 1:6) {
    MS_Table_norm = as.data.frame(cbind(MS_Table_norm, matrices[[j]][, i]))
  }
}

colnames(MS_Table_norm) = colnames(MS_Table)
rownames(MS_Table_norm) = rownames(MS_Table)

MS_Table_norm
```


## Visualization of Normalization
## Setup of Color Palette and Themes for Plots
```{r}
## Visualization of Normalization

## Setup of Color Palette and Themes for Plots
```{r}
# Hex codes for different colors:

 "#C7D5DE" # Bar title
 "#DFE3EA" # Background color
 "#617297" # Header boxes

"#A786C2"; "#8684C2"; "#617297"; "#7396AD"; "#8CABB5"; "#728083"
"#EDE6F2"; "#E7E6F2"; "#DFE2EA"; "#E2EAEE"; "#E7EEF0"; "#E2E5E6"
"#DBCEE6"; "#CFCDE6"; "#BFC6D5"; "#C6D4DE"; "#D0DDE0"; "#C5CCCD"
"#C9B6D9"; "#B7B5DA"; "#9FAAC1"; "#AABFCC"; "#BACCD2"; "#A9B3B5"
"#7E51A4"; "#504DA4"; "#485571"; "#527286"; "#628691"; "#556062"
"#54366D"; "#35336D"; "#30394B"; "#364B5A"; "#415961"; "#394041"


# Define central color palette
farben_status <- c("raw" = "#7E51A4", "normalized" = "#504DA4")
farben_status_cap <- c("Raw" = "#7E51A4", "Normalized" = "#504DA4") # For capitalized labels
farben_treatment <- c("Ctrl" = "#8CABB5", "RNAse" = "#A786C2")
farben_condition <- c("Control" = "#8CABB5", "RNase" = "#A786C2")


# Uniform theme for posters
theme_poster <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )
```


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(patchwork) # for side-by-side plots

# Function: convert to long data frame with Protein, Sample, Intensity
to_long_df <- function(df, status_label) {
  df %>%
    as.data.frame() %>%
    mutate(Protein = rownames(.)) %>%
    pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
    mutate(Status = status_label)
}

raw_long <- to_long_df(MS_Table, "raw")
norm_long <- to_long_df(MS_Table_norm, "normalized")
combined <- bind_rows(raw_long, norm_long)

combined <- combined %>%
  mutate(
    Fraction = str_extract(Sample, "Fraction\\d+"),
    Fraction_num = as.numeric(str_extract(Fraction, "\\d+")),
    Treatment = ifelse(grepl("Ctrl", Sample), "Ctrl", "RNAse")
  )

# Log transform (add 1 to avoid log(0))
combined <- combined %>%
  mutate(LogIntensity = log10(Intensity + 1))

# Only Ctrl or RNAse, here example Ctrl
combined_ctrl <- combined %>% filter(Treatment == "Ctrl")

# Prepare heatmap
plot_heatmap <- function(data, title) {
  ggplot(data, aes(x = factor(Fraction_num), y = Protein, fill = LogIntensity)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma") +
    labs(x = "Fraction", y = "Protein", fill = "log10(Intensity+1)", title = title) +
    theme_minimal() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
}

p1 <- plot_heatmap(filter(combined_ctrl, Status == "raw"), "Raw Intensities (Ctrl)")
p2 <- plot_heatmap(filter(combined_ctrl, Status == "normalized"), "Normalized Intensities (Ctrl)")

p1 + p2  + plot_layout(ncol = 2)
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Example: calculate mean values per fraction and status
prepare_summary <- function(df, status_label) {
  df %>%
    as.data.frame() %>%
    mutate(Protein = rownames(.)) %>%
    pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
    mutate(Status = status_label,
           Fraction = as.numeric(gsub("Fraction(\\d+)_.*", "\\1", Sample)),
           Treatment = ifelse(grepl("Ctrl", Sample), "Ctrl", "RNAse")) %>%
    group_by(Treatment, Fraction, Status) %>%
    summarise(mean_intensity = mean(Intensity, na.rm=TRUE), .groups="drop")
}

raw_summary <- prepare_summary(MS_Table, "raw")
norm_summary <- prepare_summary(MS_Table_norm, "normalized")

combined_summary <- bind_rows(raw_summary, norm_summary)

# Ribbon plot for raw and normalized (Ctrl and RNAse)
ggplot(combined_summary, aes(x=Fraction, y=mean_intensity, fill=Status, color=Status)) +
  geom_ribbon(aes(ymin=0, ymax=mean_intensity), alpha=0.3) +
  geom_line(size=1) +
  facet_wrap(~Treatment) +
  scale_y_log10() +
  labs(title="Protein profile across fractions: raw vs. normalized",
       x="Fraction", y="Mean intensity (log10)") +
  scale_fill_manual(values = farben_status) +
  scale_color_manual(values = farben_status) +
  theme_minimal()

```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Helper function: calculate SD per protein and fraction across replicates
calc_protein_sd <- function(data, status) {
  df <- data
  df$Protein <- rownames(df)

  df_long <- df %>%
    pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
    mutate(
      Fraction = as.numeric(gsub("Fraction(\\d+)_.*", "\\1", Sample)),
      Treatment = ifelse(grepl("Ctrl", Sample), "Ctrl", "RNAse")
    )

  # SD across replicates per protein and fraction
  df_sd <- df_long %>%
    group_by(Protein, Fraction, Treatment) %>%
    summarise(SD = sd(Intensity, na.rm = TRUE), .groups = "drop") %>%
    mutate(Status = status)

  return(df_sd)
}

# Calculate SD per protein per fraction
sd_raw <- calc_protein_sd(MS_Table, "Raw")
sd_norm <- calc_protein_sd(MS_Table_norm, "Normalized")

sd_combined <- bind_rows(sd_raw, sd_norm)

# Average SD per fraction & treatment (averaged across proteins)
sd_summary <- sd_combined %>%
  group_by(Fraction, Treatment, Status) %>%
  summarise(mean_SD = mean(SD, na.rm = TRUE), .groups = "drop")

ggplot(sd_summary, aes(x = Fraction, y = mean_SD, color = Status)) +
  geom_line(size = 1) +
  facet_wrap(~Treatment) +
  labs(title = "Average protein variance per fraction",
       x = "Fraction", y = "Mean SD across replicates") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = farben_status_cap)
```


```{r}
library(ggplot2)
library(dplyr)

# Calculate column means (raw and normalized)
frak_mittel_raw <- colMeans(MS_Table, na.rm = TRUE)
frak_mittel_norm <- colMeans(MS_Table_norm_mittel, na.rm = TRUE)

# Function to prepare a dataframe for boxplots
prepare_boxplot_df <- function(mittelwerte_vector, norm_status) {
  df <- data.frame(Sample = names(mittelwerte_vector), Value = mittelwerte_vector)
  # Extract fraction number from sample names
  df$Fraction <- sub("Fraction(\\d+)_.*", "\\1", df$Sample)
  # Set fraction order as factor (1 to 25)
  df$Fraction <- factor(df$Fraction, levels = as.character(1:25))
  # Derive condition (Control or RNase) from sample names
  df$Condition <- ifelse(grepl("Ctrl", df$Sample), "Control", "RNase")
  df$Normalization <- norm_status
  return(df)
}

# Prepare dataframes for raw and normalized data and combine
df_raw <- prepare_boxplot_df(frak_mittel_raw, "Before")
df_norm <- prepare_boxplot_df(frak_mittel_norm, "After")
df_all <- rbind(df_raw, df_norm)

# New grouping combining fraction and condition (for x-axis labels)
df_all$Group <- paste(df_all$Fraction, df_all$Condition, sep = "_")
df_all$Group <- factor(df_all$Group, levels = unique(df_all$Group))

# Boxplot: 'Before' and 'After' normalization as colors side-by-side per group
ggplot(df_all, aes(x = Group, y = Value, fill = Normalization)) +
  geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.75)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
        legend.position = "right") +
  labs(title = "Boxplots Before and After Normalization",
       x = "Fraction & Condition",
       y = "Signal Intensity",
       fill = "Status")


# Second boxplot: Compare fraction & condition with facets for normalization status
ggplot(df_all, aes(x = Fraction, y = Value, fill = Condition)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~Normalization, ncol = 2) +  # display side by side
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  scale_fill_manual(values = farben_condition) +
  labs(title = "Comparison: Before vs. After Normalization",
       x = "Fraction", y = "Intensity", fill = "Condition")
```

# For poster and template: workflow diagram with DiagrammeR, export as PNG

```{r}
# Required packages (install once, commented out)
# install.packages("DiagrammeRsvg")
# install.packages("rsvg")

# For poster and template: workflow diagram with DiagrammeR, export as PNG

# Required packages (install once, commented out)
# install.packages("DiagrammeRsvg")
# install.packages("rsvg")

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

# Create flowchart of the data cleaning process
graph <- grViz("
digraph cleaning_process {
  graph [rankdir = LR, fontsize = 10, splines = ortho]
  node [shape=box, style=filled, fontname=Helvetica, fontsize=10, penwidth=1.2, fillcolor=lightblue, color=gray50]
  edge [fontsize=9, color=gray40]

  subgraph cluster_0 {
    label = 'Initial Processing'
    style = rounded
    color = lightgrey

    A [label = '1. Raw Data\\n(150 columns)']
    B [label = '2. Split by Fractions\\n(3 columns each)']
    C [label = '3. Mean per Fraction']
  }

  subgraph cluster_1 {
    label = 'Normalization'
    style = rounded
    color = lightgrey

    D [label = '4. Compute Normalization\\n(minimize replicate deviation)']
    E [label = '5. Normalize per Replicate']
  }

  subgraph cluster_2 {
    label = 'Smoothing & Scaling'
    style = rounded
    color = lightgrey

    F [label = '6. Sliding Window\\n(3-column average)']
    G [label = '7. Scale to 100%\\nper Fraction/Replicate']
  }

  A -> B -> C -> D -> E -> F -> G
}
")

# Export diagram as SVG and convert to PNG
svg <- export_svg(graph)
rsvg_png(charToRaw(svg), file = "data_cleaning_workflow.png", width = 1200, height = 600)
```

```{r}
initial_processing = grViz("
digraph initial_processing {
  graph [rankdir = LR, fontsize = 10, splines = ortho, fontname='Arial']
  node [shape=box, style=filled, fontname='Arial', fontsize=10, penwidth=1.2, fillcolor='#EDE6F2', color=black]
  edge [fontsize=9, color=black]

  subgraph cluster_0 {
    label = 'Initial Processing'
    style = rounded
    color = lightgrey

    A [label = '1. Raw Data\\n(150 columns)']
    B [label = '2. Split by Fractions\\n(3 columns each)']
    C [label = '3. Mean per Fraction']

    A -> B -> C
  }
}
")

normalization = grViz("
digraph normalization {
  graph [rankdir = LR, fontsize = 10, splines = ortho, fontname='Arial']
  node [shape=box, style=filled, fontname='Arial', fontsize=10, penwidth=1.2, fillcolor='#EDE6F2', color=black]
  edge [fontsize=9, color=black]

  subgraph cluster_1 {
    label = 'Normalization'
    style = rounded
    color = lightgrey

    D [label = '4. Compute Normalization\\n(minimize replicate deviation)']
    E [label = '5. Normalize per Replicate']

    D -> E
  }
}
")


smoothing_scaling = grViz("
digraph smoothing_scaling {
  graph [rankdir = LR, fontsize = 10, splines = ortho, fontname='Arial']
  node [shape=box, style=filled, fontname='Arial', fontsize=10, penwidth=1.2, fillcolor='#EDE6F2', color=black]
  edge [fontsize=9, color=black]

  subgraph cluster_2 {
    label = 'Smoothing & Scaling'
    style = rounded
    color = lightgrey

    F [label = '6. Sliding Window\\n(3-column average)']
    G [label = '7. Scale to 100%\\nper Fraction/Replicate']

    F -> G
  }
}
")

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

# Example: export initial_processing
svg <- export_svg(initial_processing)
rsvg_png(charToRaw(svg), "Postervisuals/initial_processing.png", width = 3000, height = 3000)

# Normalization
svg <- export_svg(normalization)
rsvg_png(charToRaw(svg), "Postervisuals/normalization.png", width = 3000, height = 3000)

# Smoothing & Scaling
svg <- export_svg(smoothing_scaling)
rsvg_png(charToRaw(svg), "Postervisuals/smoothing_scaling.png", width = 3000, height = 3000)
```

```{r}
library(ggplot2)
library(dplyr)

# Calculate mean for each column (raw and normalized)
mean_raw <- colMeans(MS_Table, na.rm = TRUE)
mean_norm <- colMeans(MS_Table_norm, na.rm = TRUE)

# Function to extract metadata from column names
extract_meta <- function(sample_names) {
  meta_df <- data.frame(Sample = sample_names, stringsAsFactors = FALSE)
  meta_df$Fraction <- sub("Fraction(\\d+)_.*", "\\1", meta_df$Sample)
  meta_df$Fraction <- factor(meta_df$Fraction, levels = as.character(1:25))
  meta_df$Condition <- ifelse(grepl("Ctrl", meta_df$Sample), "Control", "RNase")
  meta_df$Replicate <- sub(".*_Rep(\\d+)", "\\1", meta_df$Sample)
  meta_df$Replicate <- factor(meta_df$Replicate, levels = c("1", "2", "3"))
  return(meta_df)
}

# Get metadata for raw and normalized data
meta_raw <- extract_meta(names(mean_raw))
meta_norm <- extract_meta(names(mean_norm))

# Create data frames for raw and normalized means
df_raw <- data.frame(Sample = names(mean_raw), Value = mean_raw, Normalization = "Before") %>%
  left_join(meta_raw, by = "Sample")

df_norm <- data.frame(Sample = names(mean_norm), Value = mean_norm, Normalization = "After") %>%
  left_join(meta_norm, by = "Sample")

# Combine both data frames
df_all <- bind_rows(df_raw, df_norm)

# Percent scale (normalized data is already percent, raw data scaled to percent for comparison)
df_all <- df_all %>%
  group_by(Normalization, Condition, Replicate) %>%
  mutate(Value_scaled = ifelse(Normalization == "Before", Value / max(Value, na.rm = TRUE) * 100, Value)) %>%
  ungroup()

# Set factor levels to ensure 'Before' is on the left, 'After' on the right
df_all$Normalization <- factor(df_all$Normalization, levels = c("Before", "After"))

# Plot with replicate distinction using line type and point shape
Normalization_plot <- ggplot(df_all, aes(x = Fraction, y = Value_scaled, color = Condition, group = interaction(Condition, Replicate))) +
  geom_line(aes(linetype = Replicate), alpha = 0.7) +
  geom_point(aes(shape = Replicate), size = 2) +
  facet_wrap(~Normalization, ncol = 2, scales = "free_y", strip.position = "top") + # Title on top
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.placement = "outside",
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  scale_color_manual(values = treatment_colors) +
  scale_linetype_manual(values = c("1" = "solid", "2" = "dashed", "3" = "dotted")) +
  scale_shape_manual(values = c(16, 17, 15)) +
  labs(
    title = "Mean Protein Intensities Before and After Normalization (Percent Scale)",
    x = "Fraction",
    y = "Mean Intensity (%)",
    color = "Condition",
    linetype = "Replicate",
    shape = "Replicate"
  )

print(Normalization_plot)

```

# Reproducibility

Explanation which correlation method we use:
Pearson, not Spearman, because we have many zeros — so [Pearson]{.underline}

Procedure:

```{r}
# Vector for correlations between columns 1 & 2
kor_Spalte1_Spalte2 = c()
for (i in 1:50) {
  spalten_index = ((i-1) * 3 + 1):(i * 3)  # every loop: 3 columns grouped
  teilmatrix = MS_Table_norm[,spalten_index]  # submatrix for each fraction
  
  kor_a = cor(teilmatrix[,1], teilmatrix[,2], method = 'pearson')  # Pearson correlation between column 1 & 2
  kor_Spalte1_Spalte2 = c(kor_Spalte1_Spalte2, kor_a)  # append to vector
}

kor_Spalte1_Spalte2
```


```{r}
# Vector for correlations between columns 2 & 3
kor_Spalte2_Spalte3 = c()
for (i in 1:50) {
  spalten_index = ((i-1) * 3 + 1):(i * 3)  # group 3 columns per loop
  teilmatrix = MS_Table_norm[,spalten_index]  # submatrix for each fraction
  
  kor_b = cor(teilmatrix[,2], teilmatrix[,3], method = 'pearson')  # Pearson correlation between column 2 & 3
  kor_Spalte2_Spalte3 = c(kor_Spalte2_Spalte3, kor_b)  # append to vector
}

kor_Spalte2_Spalte3

```


```{r}
# Vector for correlations between columns 1 & 3
kor_Spalte1_Spalte3 = c()
for (i in 1:50) {
  spalten_index = ((i-1) * 3 + 1):(i * 3)  # group 3 columns per loop
  teilmatrix = MS_Table_norm[,spalten_index]  # submatrix for each fraction
  
  kor_c = cor(teilmatrix[,1], teilmatrix[,3], method = 'pearson')  # Pearson correlation between column 1 & 3
  kor_Spalte1_Spalte3 = c(kor_Spalte1_Spalte3, kor_c)  # append to vector
}

kor_Spalte1_Spalte3
```


```{r}
kor_tabelle = cbind(kor_Spalte1_Spalte2, kor_Spalte2_Spalte3, kor_Spalte1_Spalte3)
rownames(kor_tabelle) = Fraktionsnamen  # assign fraction names as row names
kor_tabelle
```

## Visualization of Reproducibility

### Boxplot for the Table
```{r}
kor_long <- data.frame(
  Fraction = rep(Fraktionsnamen, 3),
  Correlation = c(kor_Spalte1_Spalte2, kor_Spalte2_Spalte3, kor_Spalte1_Spalte3),
  Pair = rep(c("Replicates 1&2", "Replicates 2&3", "Replicates 1&3"), each = 50)
)

Boxplot_Correlations <- ggplot(kor_long, aes(x = Pair, y = Correlation, fill = Pair)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Reproducibility of Replicates (Pearson Correlation)",
       x = "Replicate Pair",
       y = "Correlation Coefficient") +
  theme(legend.position = "none")

p_violin <- ggplot(kor_long, aes(x = Pair, y = Correlation, fill = Pair)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_fill_manual(values = c("#7E51A4", "#485571", "#504DA4")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Reproducibility of Replicates (Pearson Correlation)",
    y = "Correlation Coefficient",
    x = NULL  # Remove X-axis title but keep labels
  ) +
  theme(
    axis.title.y = element_text(margin = margin(r = 15)),  # space on Y-axis title
    axis.title.x = element_blank(),                         # no X-axis title
    axis.ticks.x = element_blank(),                         # no ticks on X-axis
    axis.line.x = element_blank(),                          # no bottom axis line
    legend.position = "none"
  )

print(p_violin)

# Save plot as PNG
ggsave(filename = "Postervisuals/Violinplot_Correlations.png", plot = p_violin, width = 8, height = 6)
```


### Scatterplots for all individual correlations — not part of the template
```{r}
library(stringr)
library(dplyr)
library(ggplot2)
library(tidyr)

# 1. Load data
ms = MS_Table_norm

# 2. Check
if (ncol(ms) == 0) stop("The table has no columns. Please check the CSV.")
cat("Loaded columns:\n")
print(colnames(ms))

# 3. Parse column names
info <- str_split_fixed(colnames(ms), "_", 3)
if (ncol(info) < 3) stop("Column names are not in expected format 'Fraction_Treatment_Replicate'.")

colnames(info) <- c("Fraction", "Treatment", "Replicate")
df_info <- data.frame(ColName = colnames(ms),
                      Fraction = info[, "Fraction"],
                      Treatment = info[, "Treatment"],
                      Replicate = info[, "Replicate"],
                      stringsAsFactors = FALSE)

# 4. Setup plot layout
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))  # 2x2 plots per page

# 5. For each group of Fraction + Treatment
groups <- df_info %>% group_by(Fraction, Treatment) %>% group_split()

for(group in groups) {
  if(nrow(group) < 2) next  # Need at least 2 replicates for comparison

  # All pairwise combinations of replicate columns
  combis <- combn(group$ColName, 2, simplify = FALSE)

  for(pair in combis) {
    col1 <- pair[1]
    col2 <- pair[2]
    
    x <- as.numeric(ms[[col1]])
    y <- as.numeric(ms[[col2]])
    valid <- which(!is.na(x) & !is.na(y))

    if(length(valid) >= 2) {
      plot(x[valid], y[valid],
           pch = 20, cex = 0.6,
           xlab = col1,
           ylab = col2,
           main = paste(group$Fraction[1], group$Treatment[1]))
      
      r <- cor(x[valid], y[valid])
      legend("topleft", legend = paste("R =", round(r, 2)), bty = "n")
    } else {
      plot.new()
      title(main = paste("Not enough data for", col1, "vs", col2))
    }
  }
}
```


### Example plot for median R-value

```{r}
# 1. Calculate median, ignore NAs
median_r <- median(kor_tabelle, na.rm = TRUE)

# 2. Matrix of absolute differences to median
diffs <- abs(kor_tabelle - median_r)

# 3. Index of minimal difference (NAs automatically ignored)
min_index <- which(diffs == min(diffs, na.rm = TRUE), arr.ind = TRUE)[1, , drop = FALSE]

# 4. Extract fraction and pair (fallback if names missing)
median_fraction <- rownames(kor_tabelle)[min_index[1]] %||% min_index[1]
median_pair <- colnames(kor_tabelle)[min_index[2]] %||% min_index[2]

# 5. Output results
cat("Median correlation value:", round(median_r, 3), "\n")
cat("Closest value in table:", round(kor_tabelle[min_index], 3), "\n")
cat("Plot for fraction:", median_fraction, "| pair:", median_pair, "\n")
```

#Function to plot correlation with ggplot2
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

plot_correlation <- function(data, col1, col2, color = "", save_path = "") {
  # Filter complete cases
  df <- data %>%
    select(all_of(c(col1, col2))) %>%
    mutate(Protein = rownames(data)) %>%
    pivot_longer(-Protein, names_to = "Sample", values_to = "Value") %>%
    pivot_wider(names_from = Sample, values_from = Value) %>%
    filter(complete.cases(.))
  
  # Calculate correlation
  r <- cor(df[[col1]], df[[col2]])
  r_text <- paste0("R = ", round(r, 2))
  
  # Create ggplot
  p <- ggplot(df, aes_string(x = col1, y = col2)) +
    geom_point(alpha = 0.6, color = color) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste("Scatterplot:", col1, "vs", col2),
         x = col1,
         y = col2) +
    theme_minimal() +
    annotate("text", x = min(df[[col1]], na.rm = TRUE), 
             y = max(df[[col2]], na.rm = TRUE), 
             label = r_text, hjust = 0, vjust = 1, size = 10, color = color)
  
  # Show plot
  print(p)
  
  # Save plot optionally
  if (save_path != "") {
    ggsave(filename = save_path, plot = p, width = 8, height = 8)
    cat("Plot saved at:\n", save_path, "\n")
  }
}

# Example call:
plot_correlation(MS_Table_norm, "Fraction18_Ctrl_Rep2", "Fraction18_Ctrl_Rep3", color = "black", save_path = "Postervisuals/vergleich_plot_ggplot.png")
```


Save the table as CSV (for Github upload, download in next document)
```{r}
# write.csv(MS_Table_norm, file = "MS_Table.norm.csv")
```
